---
title: "ecplore wecare ampliseq vqsr vcf"
output: html_document
editor_options: 
  chunk_output_type: console
---

started: AL07Aug2018  
last updated: AL07Aug2018

# Summary

- Import data to R  
- Make plots equivalent to bcf-stats  
- Make histograms for depth, VQSR-LOD etc
- Select thresholds for filtering  
- Filter variants:

###33,229 -> 8,745

###First, exclude genotypes:
dp < 10  
dp > 10,000  
gq < 20  

###Then exclude variants:
Call rate < 80% (~1/3)  
Failed VQSR (~1/3)  

# start_section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r set_width, echo=F}

options(width = 999)
# https://stackoverflow.com/questions/36845178/width-of-r-code-chunk-output-in-rmarkdown-files-knitr-ed-to-html

```

```{r start_section}

Sys.time()
rm(list=ls())
graphics.off()

library(knitr)
library(vcfR) # the main work-horse
library(stringr) # for word
library(tidyr) # for separate

base_folder="/Users/alexey/Documents/wecare/ampliseq/a07_vqsr_vcf"
opts_knit$set(root.dir = base_folder)
options(stringsAsFactors = FALSE)

```

# read_vcf  

**Strange NAs in well covered positions (see NA-s in 1_3669317 and 1_3669318)**  
However, NA rate is ~7%, so this should affect only a small proportion of data  

```{r read_vcf}

# Source VCF
vcf_file=paste(base_folder,"wecare_ampliseq_vqsr.vcf",sep="/")

# read vcf to vcfR object
vcfr <- read.vcfR(vcf_file)
  
# get data from header and fixed columns
meta_fix <- vcfR2tidy(vcfr, info_only=T)
    
# data frame with meta-information from vcf header
meta.df <- meta_fix$meta
dim(meta.df)
meta.df

# data frame with fixed columns (including parsed INFO, convert tibble to data-frame)
fixed.df <- as.data.frame(meta_fix$fix)
dim(fixed.df)
colnames(fixed.df)
fixed.df[15:20,1:7]

# extract genotypes matrix (with numeric representation of alleles)
gt_num.mx <- extract.gt(vcfr) # original numeric codes: 0/1, 1/1 etc
dim(gt_num.mx)
gt_num.mx[15:20,1:5]
sum(is.na(gt_num.mx))
sum(!is.na(gt_num.mx))
sum(is.na(gt_num.mx))/sum(!is.na(gt_num.mx))

# extract genotypes matrix (keep alleles as ACGT)
gt_chr.mx <- extract.gt(vcfr, return.alleles = TRUE) # encoded as A/A, T/G etc
dim(gt_chr.mx)
gt_chr.mx[15:20,1:5]
NA -> gt_chr.mx[gt_chr.mx=="."]
gt_chr.mx[15:20,1:5]

# extract DP, GQ and AD matrices
dp.mx <- extract.gt(vcfr, element = "DP", as.numeric = TRUE)
dim(dp.mx)
dp.mx[15:20,1:5]

gq.mx <- extract.gt(vcfr, element = "GQ", as.numeric = TRUE)
dim(gq.mx)
gq.mx[15:20,1:5]

ad.mx <- extract.gt(vcfr, element = "AD")
dim(ad.mx)
ad.mx[15:20,1:5]

# Clean-up
rm(vcf_file, vcfr, meta_fix)

```

# add rownames to fixed.df

```{r add_rownames}

# Generate row names
key <- paste(fixed.df$CHROM, fixed.df$POS, sep="_")

# Explore and check identity with genotypes matrix
key[1:5]
rownames(gt_num.mx)[1:5]
sum(key != rownames(gt_num.mx))

# Assign the rownames
key -> rownames(fixed.df)

# Clean-up
rm(key)

```

# explore QC metrics

```{r explore_qc_metrics}

# NA rate
sum(is.na(gt_num.mx))
sum(is.na(gt_num.mx))/(nrow(gt_num.mx)*ncol(gt_num.mx))

# VQSR
sum(is.na(fixed.df$FILTER))
sort(table(fixed.df$FILTER), decreasing = T)
hist(fixed.df$VQSLOD)
hist(fixed.df$VQSLOD[fixed.df$VQSLOD > -50])

# DP (cumulative depth per variant)
sum(is.na(fixed.df$DP))
hist(fixed.df$DP, lab=T, ylim=c(0,35000))$breaks

# dp.mx (individual depth per sample)
sum(is.na(dp.mx))
hist(dp.mx, lab=T)$breaks
sum(dp.mx < 10, na.rm=T)/sum(!is.na(dp.mx))
sum(dp.mx > 10000, na.rm=T)/sum(!is.na(dp.mx))

# gq.mx (individual genotype quality per sample)
sum(is.na(gq.mx))
hist(gq.mx, lab=T)$breaks
sum(gq.mx < 20, na.rm=T)/sum(!is.na(gq.mx))

```

# filter variants

33,229 -> 8,745

###First, exclude genotypes:
dp < 10
dp > 10,000
gq < 20

###Then exclude variants:
Call rate < 80%
Failed VQSR

###To consider:
DP > 1M (?)

##filter_genotypes

```{r filter_genotypes}

# Initialise matrix for filtered variants
gt_num_filt.mx <- gt_num.mx

# Check NA rate
sum(is.na(gt_num_filt.mx))
sum(is.na(gt_num_filt.mx))/(nrow(gt_num_filt.mx)*ncol(gt_num_filt.mx))

# Filter genotypes
NA -> gt_num_filt.mx[ dp.mx < 10 ]
NA -> gt_num_filt.mx[ dp.mx > 10000 ]
NA -> gt_num_filt.mx[ gq.mx < 20 ]

# Check NA rate
sum(is.na(gt_num_filt.mx))
sum(is.na(gt_num_filt.mx))/(nrow(gt_num_filt.mx)*ncol(gt_num_filt.mx))

```

##calculate_call_rates

```{r calculate_call_rates}

# Function to get call rate
call_rate.udf <- function(x){sum(!is.na(x))/length(x)}

# Get call rates
call_rates <- apply(gt_num_filt.mx,1,call_rate.udf)
hist(call_rates)

# Clean-up
rm(call_rate.udf)

```

##filter_varaints

```{r filter_varaints}

# Check intial num of variants
dim(gt_num_filt.mx)

# Prepare variants filter
vqsr_passed <- fixed.df$FILTER == "PASS"
sum(vqsr_passed)

call_rate_passed <- call_rates >= 0.8
sum(call_rate_passed)

selected <- vqsr_passed & call_rate_passed
sum(selected)

# Apply variants filter to all tables
gt_num_filt.mx <- gt_num_filt.mx[selected,]
gt_chr_filt.mx <- gt_chr.mx[selected,]
ad_filt.mx <- ad.mx[selected,]
dp_filt.mx <- dp.mx[selected,]
gq_filt.mx <- gq.mx[selected,]
fixed_filt.df <- fixed.df[selected,]

# Clean-up
rm(vqsr_passed, call_rate_passed, call_rates, selected)

```

##explore_result

```{r explore_result}

# Dimentions of tables
dim(gt_num_filt.mx)
dim(gt_chr_filt.mx)
dim(ad_filt.mx)
dim(dp_filt.mx)
dim(gq_filt.mx)
dim(fixed_filt.df)

# NA rate
sum(is.na(gt_num_filt.mx))
sum(is.na(gt_num_filt.mx))/(nrow(gt_num_filt.mx)*ncol(gt_num_filt.mx))

# Selected histograms
hist(fixed_filt.df$DP)
hist(fixed_filt.df$VQSLOD)

```

# Export retained variants

```{r export_retained_variants}

# Prepare output data
targets.df <- fixed_filt.df[,c("CHROM","POS")]

# Save to text file
out_file=paste(base_folder, "targets.txt", sep="/")
write.table(targets.df, file=out_file, sep="\t", quote=F, row.names=F, col.names=F)

# Clean-up
rm(targets.df, out_file)

```

## More exploring/filtering?
make histograms for each metric; plot 2/3 metrics against each other, split by PASS, etc

# save_results

```{r save_results}

save.image(paste(base_folder, "explore_wecare_ampliseq_vqsr_vcf.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
